Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-24 paper.Rnw
FEP with FS/FIRST

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx, graphicx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}   % keep floats in their place.
\usepackage{authblk}             % Listing Author affiliations
\usepackage{rotating}            % rotating figures
\usepackage[hypcap]{caption}
\usepackage{lineno}

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\multiatlas}{multi-atlas}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\FEPdataset}{SZ-FEP}
\newcommand{\fsl}{FSL FIRST}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}
\newcommand{\prsnr}{Pruessner}
\newcommand{\pseg}{\prsnr-protocol manual segmentation}
\newcommand{\psegs}{\prsnr-protocol manual segmentations}
\newcommand{\snt}{SNT}

@@

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas{} approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} 
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
\multiatlas{} approaches. Thus, our method acts as an accurate \multiatlas{}
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template
library, formed from a subset of target images, via transformations estimated by
non-linear image registration. The resultant segmentations are then propagated
to each target image and fused using a label fusion method.  To explore the
effect of atlases and templates library size, registation and label
fusion method, we conduct a 10-fold Monte Carlo cross-validation of \mb{} of whole
hippocampal segmentation on 60 ADNI subjects, manually segmented using the
\prsnr{}-protocol, over a range of parameter settings. Using the best settings
found, we then conduct a leave-one-out cross-validation (LOOCV) of hippocampal
subfield segmentation using five high-resolution manually segmented atlases.
Two final experiments assess \mb{} when applied to first episode psychosis and
Alzheimer's disease populations, and \mb{} segmentations are compared with
existing automated methods (\fsl{}, \freesurfer{}, \maper{}) and biases are
explored.\\
\textbf{Results:} Using 9 atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of \Sexpr{round(mb.kappa.best, digits=3)} 
(to \psegs{}) over 10-folds of Monte Carlo % TODO: update cross-validation on 60 subjects, and shows
significantly lower variability in DSC than \multiatlas{} segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1;
CA4/dentate % TODO: update
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subfields with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt{})
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.\\ \textbf{Conclusion:} We demonstrate
that \mb{} produces accurate hippocampal segmentations using only 5 atlases over
different hippocampal definitions, disease populations, and acquisition types,
as well as showing that accurate identification of the hippocampal subfields is
possible.
\end{abstract}

@@

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer{} package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were labelled on highly-anistropic
($0.380mm$ in-plane resolution, $0.8mm$ coronal slice thickness) T2-weighted MR
images acquired from post-mortem medial temporal lobe samples. Using nonlinear
registration guided by manually derived hippocampus masks and specific
landmarks, the authors demonstrate accurate parcellation of hippocampal
subfields in unlabelled MRI volumes.

In this paper we have describe a thorough validation of the \mb{} algorithm for
segmentation of the hippocampus and its subfields. First, we address the
feasibility of whole hippocampus segmentation with a limited number of input
atlases \citep{Chakravarty2013} by performing a multi-fold validation experiment using a
subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset
manually segmented using the \prsnr{}-protocol, over a
range of atlas and template library sizes, registration and label fusion
methods. We then perform a leave-one-out validation to determine if
hippocampal subfields can be accurately identified, using the best parameters
discovered in the first experiment.  To ensure that we have not overfit our
parameters to the aging or neurodegenerative brain, we also apply \mb{} to a
dataset of individuals suffering from first episode psychosis.  Finally, we
validate our algorithm using all of the data available in the \adnidataset{}
sample and compare our segmentations to other popular segmentation algorithms.

@@

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic \multiatlas
segmentation.  Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls).  The template library images are automatically
labelled by each of the atlases via label propagation.  Basic \multiatlas
segmentation is then conducted using the template library to segment the entire
set of target images (including the target images used in the construction of
the template library).  Since each template library image has multiple labels
(one from each atlas), the final number of labels to be fused for each target
may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic \multiatlas{} segmentation and
\mb{} segmentation. In multi-atlas segmentation, manual labels from atlas images
are warped (propagated) into subject space by applying the transformations
estimated from non-linear image registration. The resulting candidate labels
from all atlas images are then fused to create a final segmentation. In \mb{}
segmentation, a template library is created by sampling (either randomly or
representatively) from the subject images. Atlas labels are propagated to all
template images and then to each subject image (including those used in the template
library). The candidate the labels for a subject are then fused into a final
segmentation.  \label{fig:MAGeT}}
\end{figure}

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb{} algorithm. The first two experiments assess the
validity of \mb{} using cross-validation designs. Experiment 1 investigates the
accuracy of \mb{} whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb{} algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compare \mb{} segmentations with those of
well-known automated methods and assessed segmentation bias.

@@

In the Supplementary Materials we have replicated this experiment using the \snt
semi-automated labels included with the ADNI dataset.

\paragraph{\adnidataset{} dataset} 
Data used in the preparation of this article were obtained from the Alzheimer's
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public-private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer's disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

The Principal Investigator of this initiative is Michael W. Weiner, MD, VA
Medical Center and University of California San Francisco. ADNI is the result
of efforts of many co- investigators from a broad range of academic institutions
and private corporations, and subjects have been recruited from over 50 sites
across the U.S. and Canada. The initial goal of ADNI was to recruit 800 subjects
but ADNI has been followed by ADNI-GO and ADNI-2. To date these three protocols
have recruited over 1500 adults, ages 55 to 90, to participate in the research,
consisting of cognitively normal (CN) older individuals, people with early or
late MCI, and people with early AD. The follow up duration of each group is
specified in the protocols for ADNI-1, ADNI-2 and ADNI-GO. Subjects originally
recruited for ADNI-1 and ADNI-GO had the option to be followed in ADNI-2. For
up-to-date information, see www.adni-info.org.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset{}" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

Before registration, all images underwent preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In this experiment we
compared two non-linear image registration methods:

@@

\begin{figure}
  \begin{centering}
    % TODO: fix schematic
    \includegraphics[width=\textwidth]{figure/WinterburnSubfieldLOOCVSchematic}
  \end{centering}
  \caption{A schematic illustration of the experimental set up of the
  leave-one-out cross-validation (LOOCV) subfield experiment.}
  \label{fig:LOOCV-schematic}
\end{figure}

@@

To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases to derive 
whole hippocampal segmentations of a dataset of patients with Schizophrenia. The
resulting segmentations are assessed for quality by comparison with expert
manual segmentations.

\paragraph{\FEPdataset{} dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 14 to 35 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

\paragraph{Experiment details} 
\mb{} is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2) ignoring subfield delineations.  All images from the
\FEPdataset{} dataset are segmented by \mb{}.  The optimal size of template library,
registration method, and label fusion method found in Experiment 1 are used.

@@

In addition to a mean increase in similarity score over \multiatlas{}-based
segmentation, \mb{} also shows more consistency in similarity scores across all
subjects and validation folds (Figure \ref{ADNI1-xval-variability}). A
template library of at least 13 images is sufficient to show significant ($p <
0.05$) decrease in variance for all sizes of atlas library tested (1-9 images).

We have omitted results obtained when using an even number of atlases or
templates since with this configuration we found significantly decreased
performance. We believe this is as a result of an inherent bias in the majority
vote fusion method used (see Discussion).

@@

@@

Figure \ref{fig:WAval-vol-boxplot}) shows the mean relative percent error in volume, 
with respect to the full-resolution Winterburn atlas, of each hippocampal subfield
of the resampled Winterburn labels, and of the \mb{} segmentations of the resampled
Winterburn atlas images and T1 BRAVO Winterburn subject acquisitions.  A
paired-samples t-test was conducted to compare the distribution of relative
percent error in the \mb{} segmentation volumes and the resampled segmentation
volumes. There was no significant difference found for the CA1, Subiculum, and
CA2/CA3 subregions. The CA4/DG subregion shows significantly lower percent error 
  ($M_{res}=$\Sexpr{wa.s$m_res[wa.s$region=='CA4/DG']},
  $SD_{res}=$\Sexpr{wa.s$sd_res[wa.s$region=='CA4/DG']};
  $M_{mb-bravo}=$\Sexpr{wa.s$m_bravo[wa.s$region=='CA4/DG']},    
  $SD_{mb-bravo}=$\Sexpr{wa.s$sd_bravo[wa.s$region=='CA4/DG']};    
  $M_{mb-res}=$\Sexpr{wa.s$m_mb_res[wa.s$region=='CA4/DG']},    
  $SD_{mb-res}=$\Sexpr{wa.s$sd_mb_res[wa.s$region=='CA4/DG']};    
  $p < 0.001$), and the SR/SL/SM shows significantly different percent error in the opposite
direction 
  ($M_{res}=$\Sexpr{wa.s$m_res[wa.s$region=='SR/SL/SM']},
  $SD_{res}=$\Sexpr{wa.s$sd_res[wa.s$region=='SR/SL/SM']};
  $M_{mb-bravo}=$\Sexpr{wa.s$m_bravo[wa.s$region=='SR/SL/SM']},    
  $SD_{mb-bravo}=$\Sexpr{wa.s$sd_bravo[wa.s$region=='SR/SL/SM']};    
  $M_{mb-res}=$\Sexpr{wa.s$m_mb_res[wa.s$region=='SR/SL/SM']},    
  $SD_{mb-res}=$\Sexpr{wa.s$sd_mb_res[wa.s$region=='SR/SL/SM']};    
  $p < 0.001$). Bland-Altman plots comparing \mb{} to subsampled segmentation
volume reveals a proportional bias towards underestimating volumes in all
subregions, and a noticeable fixed bias in the CA4/DG and SR/SL/SM subregions.

Figure \ref{fig:subfield-montage} shows slices subfield segmentations for a
single subject for qualitative inspection.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: compare resampling
\begin{figure}[h!]
<<fig:WAval-vol-boxplot,cache=T,fig.width=6,fig.height=5>>=

wa.xval.error.melted = melt(wa.xval.error,id.vars=c("file", "region"))
wa.xval.error.melted = rename(wa.xval.error.melted, c("variable"="measure", "value"="percenterr"))
levels(wa.xval.error.melted$measure) <- list("Subsampled atlas labels"="resampled",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, lrl, geometry, ctable}
\usepackage{titlesec, siunitx, graphicx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}   % keep floats in their place.
\usepackage{authblk}             % Listing Author affiliations
\usepackage{rotating}            % rotating figures
\usepackage[hypcap]{caption}
\usepackage{lineno}

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\multiatlas}{multi-atlas}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\fsl}{FSL FIRST}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}
\newcommand{\prsnr}{Pruessner}
\newcommand{\pseg}{\prsnr-protocol manual segmentation}
\newcommand{\psegs}{\prsnr-protocol manual segmentations}
\newcommand{\snt}{SNT}

@@

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas{} approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} 
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
\multiatlas{} approaches. Thus, our method acts as an accurate \multiatlas{}
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template
library, formed from a subset of target images, via transformations estimated by
nonlinear image registration. The resultant segmentations are then propagated
to each target image and fused using a label fusion method.  To explore the
effect of atlases and templates library size, registation and label
fusion method, we conduct a 10-fold Monte Carlo cross-validation of \mb{} of whole
hippocampal segmentation on 60 ADNI subjects, manually segmented using the
\prsnr{}-protocol, over a range of parameter settings. Using the best settings
found, we then conduct a leave-one-out cross-validation (LOOCV) of hippocampal
subfield segmentation using five high-resolution manually segmented atlases.
Two final experiments assess \mb{} when applied to first episode psychosis and
Alzheimer's disease populations, and \mb{} segmentations are compared with
existing automated methods (\fsl{}, \freesurfer{}, \maper{}) and biases are
explored.\\
\textbf{Results:} Using 9 atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of \Sexpr{round(mb.kappa.best, digits=3)} 
(to \psegs{}) over 10-folds of Monte Carlo % TODO: update cross-validation on 60 subjects, and shows
significantly lower variability in DSC than \multiatlas{} segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1;
CA4/dentate % TODO: update
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subfields with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt{})
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.\\ \textbf{Conclusion:} We demonstrate
that \mb{} produces accurate hippocampal segmentations using only 5 atlases over
different hippocampal definitions, disease populations, and acquisition types,
as well as showing that accurate identification of the hippocampal subfields is
possible.
\end{abstract}

@@

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1- or T2-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer{} package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were manually labelled on
high-resolution (either $0.2mm^3$ isotropic or $0.2mm \times 0.3mm \times 0.2mm$
resolution voxels) T2-weighted MR images acquired from five post-mortem medial
temporal lobe samples. Using nonlinear registration guided by shape-based models
of the subfield segmentations, and manually derived hippocampus masks of the
target images, the authors demonstrate accurate parcellation of hippocampal
subfields in clinical 3T T1-weighted MRI volumes. This method is not fully
automatic as it requires whole hippocampal parcellation of the target images.
%TODO: ref other Yush paper

In this paper we have describe a thorough validation of the \mb{} algorithm for
the automatic segmentation of the hippocampus and its subfields. First, we
address the feasibility of whole hippocampus segmentation with a limited number
of input atlases \citep{Chakravarty2013} by performing a multi-fold validation
experiment using a subset of the Alzheimer's Disease Neuroimaging Initiative
(ADNI) dataset manually segmented using the \prsnr{}-protocol, over a range of
atlas and template library sizes, registration and label fusion methods. We then
perform a leave-one-out validation to determine if hippocampal subfields can be
accurately identified, using the best parameters discovered in the first
experiment.  To ensure that we have not overfit our parameters to the aging or
neurodegenerative brain, we also apply \mb{} to a dataset of individuals
suffering from first episode psychosis.  Finally, we validate our algorithm
using all of the data available in the \adnidataset{} sample and compare our
segmentations to other popular segmentation algorithms.

@@

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic \multiatlas
segmentation.  Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls).  The template library images are automatically
labelled by each of the atlases via label propagation. Effectively, basic
\multiatlas{} segmentation is then conducted using the template library to
segment the entire set of target images (including the target images used in the
construction of the template library). Since each template library image has
multiple labels (one from each atlas), the final number of labels to be fused
for each target may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic \multiatlas{} segmentation and
\mb{} segmentation. In multi-atlas segmentation, manual labels from atlas images
are warped (propagated) into subject space by applying the transformations
estimated from nonlinear image registration. The resulting candidate labels
from all atlas images are then fused to create a final segmentation. In \mb{}
segmentation, a template library is created by sampling (either randomly or
representatively) from the subject images. Atlas labels are propagated to all
template images and then to each subject image (including those used in the template
library). The candidate the labels for a subject are then fused into a final
segmentation.  \label{fig:MAGeT}}
\end{figure}

@@

The following section describes experiments conducted to assess the
segmentation quality of the \mb{} algorithm. The first two experiments assess the
validity of \mb{} using cross-validation designs. Experiment 1 investigates the
accuracy of \mb{} whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb{} algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compare \mb{} segmentations with those of
well-known automated methods and assessed segmentation bias.

@@

In the Supplementary Materials we have replicated this experiment using the
\snt{} semi-automated labels included with the ADNI dataset.

\paragraph{\adnidataset{} dataset} 
Data used in the preparation of this article were obtained from the Alzheimer's
Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public-private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer's disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

The Principal Investigator of this initiative is Michael W. Weiner, MD, VA
Medical Center and University of California San Francisco. ADNI is the result
of efforts of many co-investigators from a broad range of academic institutions
and private corporations, and subjects have been recruited from over 50 sites
across the U.S. and Canada. The initial goal of ADNI was to recruit 800 subjects
but ADNI has been followed by ADNI-GO and ADNI-2. To date these three protocols
have recruited over 1500 adults, ages 55 to 90, to participate in the research,
consisting of cognitively normal (CN) older individuals, people with early or
late MCI, and people with early AD. The follow up duration of each group is
specified in the protocols for ADNI-1, ADNI-2 and ADNI-GO. Subjects originally
recruited for ADNI-1 and ADNI-GO had the option to be followed in ADNI-2. For
up-to-date information, see www.adni-info.org.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset{}" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.usc.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

Before registration, all images underwent preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In this experiment we
compared two nonlinear image registration methods:

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/WinterburnSubfieldLOOCVSchematic}
  \end{centering}
  \caption{A schematic illustration of the experimental set up of the
  leave-one-out cross-validation (LOOCV) subfield experiment.}
  \label{fig:LOOCV-schematic}
\end{figure}

@@

To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases to derive 
whole hippocampal segmentations of a dataset of patients with schizophrenia. The
resulting segmentations are assessed for quality by comparison with expert
manual segmentations.

\paragraph{First Episode Schizophrenia Dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 14 to 35 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

\paragraph{Experiment details} 
\mb{} is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2) ignoring subfield delineations.  All images from the
first episode dataset are segmented by \mb{}.  The optimal size of template
library, registration method, and label fusion method found in Experiment 1 are
used.

@@

In addition to a mean increase in similarity score over \multiatlas{}-based
segmentation, \mb{} also shows more consistency in similarity scores across all
subjects and validation folds (Figure \ref{fig:ADNI1-xval-variability}). A
template library of at least 13 images is sufficient to show significant ($p <
0.05$) decrease in variance for all sizes of atlas library tested (1-9 images).

We find similar behaviour with respect to optimal parameter settings and
increased consistency of \mb{} segmentations in the replication of this
experiment (Experiment 5, Supplementary Materials) where atlases use a different
hippocampal definition is used. This strongly suggests that these results are
independant of the segmentation protocol used and are, instead, features of the
\mb{} algorithm.

We have omitted results obtained when using an even number of atlases or
templates since with these configurations we found significantly decreased
performance. We believe this is results from an inherent bias in the majority
vote fusion method used (see Discussion).

@@

fmt_tstat = function(a, b) {
  test = t.test(a,b, paired=T)
  p = sprintf("%0.3f", max(0.001,round(test$p.value,3)))
  t = sprintf("%0.3f", round(test$statistic,3))
  paste(t," (",p,")")
}

wa.p = ddply(wa.xval.casted, c("region"), function (df) {
  with(df, {
       data.frame(
         res = fmt_tstat(gold_0.3mm, gold_0.9mm),
         bra = fmt_tstat(gold_0.3mm, mb_bravo),
         mbr = fmt_tstat(gold_0.3mm, mb_0.9mm)
       )})
  })

wa.p = rename(wa.p, c(
  'region'='Subregion','res'='Resampled',
  'bra'='MAGeT on BRAVO','mbr'='MAGeT on Resampled'))

@@

Figure \ref{fig:WAval-vol-boxplot} shows the mean relative percent error in volume, 
with respect to the full-resolution Winterburn atlas, of each hippocampal subfield
of the resampled Winterburn labels, and of the \mb{} segmentations of the resampled
Winterburn atlas images and T1 BRAVO Winterburn subject acquisitions.  A
paired-samples t-test was conducted to compare the distribution of relative
percent error in the \mb{} segmentation volumes and the resampled segmentation
volumes. There was no significant difference found for the CA1, Subiculum, and
CA2/CA3 subregions. The CA4/DG subregion shows significantly lower percent error 
  ($M_{resampled}=$\Sexpr{wa.s$m_res[wa.s$region=='CA4/DG']},
  $SD_{resampled}=$\Sexpr{wa.s$sd_res[wa.s$region=='CA4/DG']};
  $M_{MAGeT(BRAVO)}=$\Sexpr{wa.s$m_bravo[wa.s$region=='CA4/DG']},    
  $SD_{MAGeT(BRAVO)}=$\Sexpr{wa.s$sd_bravo[wa.s$region=='CA4/DG']};    
  $M_{MAGeT(resampled)}=$\Sexpr{wa.s$m_mb_res[wa.s$region=='CA4/DG']},    
  $SD_{MAGeT(resampled)}=$\Sexpr{wa.s$sd_mb_res[wa.s$region=='CA4/DG']};    
  $p < 0.001$), and the SR/SL/SM shows significantly different percent error in the opposite
direction 
  ($M_{resampled}=$\Sexpr{wa.s$m_res[wa.s$region=='SR/SL/SM']},
  $SD_{resampled}=$\Sexpr{wa.s$sd_res[wa.s$region=='SR/SL/SM']};
  $M_{MAGeT(BRAVO)}=$\Sexpr{wa.s$m_bravo[wa.s$region=='SR/SL/SM']},    
  $SD_{MAGeT(BRAVO)}=$\Sexpr{wa.s$sd_bravo[wa.s$region=='SR/SL/SM']};    
  $M_{MAGeT(resampled)}=$\Sexpr{wa.s$m_mb_res[wa.s$region=='SR/SL/SM']},    
  $SD_{MAGeT(resampled)}=$\Sexpr{wa.s$sd_mb_res[wa.s$region=='SR/SL/SM']};    
  $p < 0.001$). Bland-Altman plots comparing \mb{} to subsampled segmentation
volume reveals a proportional bias towards underestimating volumes in all
subregions, and prominant fixed biases in the volumes of the CA4/DG and SR/SL/SM subregions 
congruent with significant differences seen in percent error in volume
associated with subsampling alone.

Figure \ref{fig:subfield-montage} shows slices subfield segmentations for a
single subject for qualitative inspection.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WAxval - t.test comparing Manual volumes and Computed volumes of subregions 
%          across folds
<<tab:WAxval-t-test-manual-computed, echo=F, results="asis",cache=FALSE>>=
latex(wa.p, file="", size="scriptsize",landscape=FALSE, ctable=TRUE,
    caption="\\textbf{Paired-sample t-tests comparing manually segmented
             Winterburn atlas subregion volumes and computed volumes (reported
             as $t (p)$). }", 
    col.just=c("l","r","r","r"),
    caption.loc = c("bottom"), rowname=NULL, 
    label="tab:WAxval-t-test-manual-computed")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% % Winterburn-XVAL: boxplot showing raw volumes of all methods and all subjects
%% \begin{figure}[h!]
%% <<fig:WAval-volumes,cache=T,fig.width=6,fig.height=5>>=
%% m=melt(wa.xval.casted, id.vars=c('region','file'))
%% levels(m$variable) <- list("Original atlas images"="gold_0.3mm",
%%                            "Subsampled atlas images"="gold_0.9mm",
%%                            "MAGeT on subsampled atlas images" = "mb_0.9mm",
%%                            "MAGeT on T1 BRAVO images"="mb_bravo")
%% m = rename(m, c(
%%   'region'='Subregion','value'='Volume',
%%   'variable'='Method'))
%% 
%% ggplot(m,aes(y=Volume,x=Subregion,colour=Method))  + 
%%   geom_boxplot() + 
%%   geom_point(position=position_dodge(width=0.75)) +
%%   ylab(expression('Volume ' (mm^3))) + 
%%   theme(legend.direction = "horizontal", legend.position = "bottom")
%% @
%%   \caption{{\bf Manual and computed unilateral hippocampal subfield
%% volumes.}\label{fig:WAval-volumes}}
%% \end{figure}
%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
